---
output: github_document
bibliography: vignettes/citations.bib
csl: vignettes/apa.csl
link-citations: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# hmetad <a href="https://metacoglab.github.io/hmetad/"><img src="man/figures/logo.svg" align="right" height="138" alt="hmetad website" /></a>

<!-- badges: start -->
[![R-CMD-check](https://github.com/metacoglab/hmetad/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/metacoglab/hmetad/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/hmetad)](https://CRAN.R-project.org/package=hmetad)
[![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/grand-total/hmetad?color=blue)](https://r-pkg.org/pkg/hmetad)
<!-- badges: end -->

The `hmetad` package is designed to fit the meta-d’ model for confidence ratings [@maniscalco2012; @maniscalco2014]. The `hmetad` package uses a Bayesian modeling approach, building on and superseding previous development of the [Hmeta-d toolbox](https://github.com/metacoglab/HMeta-d) [@fleming2017]. A key advance is implementation as a custom family in the [brms](https://paulbuerkner.com/brms/) package, which itself provides a friendly interface to the probabilistic programming language [Stan](https://mc-stan.org).

This provides major benefits:

  - Model designs can be specified as simple `R` formulas
  - Support for complex model designs (e.g., multilevel models,
    distributional models, multivariate models)
  - Interfaces to other packages surrounding `brms` (e.g., `tidybayes`, `ggdist`, `bayesplot`, `loo`, `posterior`, `bridgesampling`, `bayestestR`)
  - Computation of model-implied quantities (e.g., mean confidence, type 1 and type 2 receiver operating characteristic curves)
  - Derivation of novel metacognitive bias metrics, and well as established metrics of metacognitive efficiency
  - Increased sampling efficiency and better convergence diagnostics



## Installation
`hmetad` is available via CRAN and can be installed using:
``` {r install, eval=FALSE}
install.packages("hmetad")
```

Alternatively, you can install the development version of `hmetad` from [GitHub](https://github.com/metacoglab/hmetad) with:
``` {r install_dev, eval=FALSE}
# install.packages("pak")
pak::pak("metacoglab/hmetad")
```

## Quick setup
Let's say you have some data from a binary decision task with ordinal confidence ratings:
```{r data, echo=FALSE, message=FALSE, warning=FALSE}
library(hmetad)
library(dplyr)
d <- sim_metad(N_trials = 1000) |>
  ungroup() |>
  select(trial:confidence) |>
  slice_sample(prop = 1) |>
  mutate(trial = row_number())
d
```

You can fit an intercepts-only meta-d' model using `fit_metad`:
```{r example, results=FALSE, warning=FALSE, message=FALSE}
library(hmetad)

m <- fit_metad(N ~ 1,
  data = d,
  prior = prior(normal(0, 1), class = Intercept) +
    set_prior(
      "normal(0, 1)",
      class = c(
        "dprime", "c",
        "metac2zero1diff", "metac2zero2diff", "metac2zero3diff",
        "metac2one1diff", "metac2one2diff", "metac2one3diff"
      )
    )
)
```

```{r summary, echo=FALSE, warning=FALSE, message=FALSE}
m
```


Now let's say you have a more complicated design, such as a within-participant manipulation:
```{r data2, echo=FALSE}
d <- sim_metad_participant_condition(N_participants = 25) |>
  group_by(participant, condition) |>
  select(participant:confidence) |>
  slice_sample(prop = 1) |>
  mutate(trial = row_number())
d
```

To account for the repeated measures in this design, you can simply adjust the formula to include participant-level effects:
```{r example2, results=FALSE, warning=FALSE, message=FALSE}
m <- fit_metad(
  bf(
    N ~ condition + (condition | participant),
    dprime + c +
      metac2zero1diff + metac2zero2diff + metac2zero3diff +
      metac2one1diff + metac2one2diff + metac2one3diff ~
      condition + (condition | participant)
  ),
  data = d, init = "0",
  prior = prior(normal(0, 1)) +
    set_prior("normal(0, 1)",
      dpar = c(
        "dprime", "c",
        "metac2zero1diff", "metac2zero2diff", "metac2zero3diff",
        "metac2one1diff", "metac2one2diff", "metac2one3diff"
      )
    )
)
```

```{r summary2, echo=FALSE, warning=FALSE, message=FALSE}
m
```

## References
